Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2021-10-22 analysis/paper/paper.Rmd
reduce code in paper

<!-- separated and mixed treatments -->
The difference in incorporation between the size categories resulted in a change
in size ratios between the original starch solutions and the extracted samples.
Large potato granules (> 20 $\mu$m) were most affected, with a 
`r paste0(abs(signif(size_diff$l[4]), 3), "%")` 
decrease in relative abundance in the potato-only treatment, and a
`r paste0(abs(signif(size_diff$l[1]), 3), "%")` 
decrease in mixed treatments. Medium granules increased in relative abundance 
across all samples, while small granules decreased in wheat treatments and 
increased in potato treatments 
(Figure \@ref(fig:ratio-plots)).

@@

<!-- Big summary/important statement for the science -->
Here, we have provided a method for exploring the incorporation of dietary 
starches into the mineral matrix of a dental calculus biofilm model. Our results show
that a very low proportion of the starches exposed to the biofilm during growth are
retained in the mineral matrix, and that the size of the starch granules
may affect the likelihood of incorporation. The proportions of starch granules 
(of all sizes) present in the extracted samples were similar across all treatments
(`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`), 
despite large differences in absolute granule counts between wheat 
(mean = `r format(sol_comb_out$total[5], big.mark = ",")`) 
and potato
(mean = `r format(sol_comb_out$total[4], big.mark = ",")`)
solutions.  
The absolute counts, however, differed more visibly between treatments and was
proportional with the total count of granules in the treatment solutions. Wheat 
and mixed solutions had the highest absolute mean count of starch granules, and 
also had the highest absolute mean count of starch granules extracted from the 
dental calculus
(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
This suggests that the starches that are more frequently consumed will be present
in higher quantities in the dental calculus, at least prior to inhumation and 
degradation in the burial environment. 
Despite the low proportion of granules recovered from the model calculus 
(`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`),
the absolute counts were still substantially greater than counts recovered from 
archaeological remains 
[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossilContent2010], <!-- although this study extrapolated counts to full sample content, which is
rarely done in archaeological studies -->which could in part be due to the 
lack of oral amylase activity in our model.

>>>>

<!-- separated and mixed treatments -->
The difference in incorporation between the size categories resulted in a change
in size ratios between the original starch solutions and the extracted samples.
Large potato granules (> 20 $\mu$m) were most affected, with a 
`r paste0(abs(signif(size_diff$l[4], 3)), "%")` 
decrease in relative abundance in the potato-only treatment, and a
`r paste0(abs(signif(size_diff$l[1], 3)), "%")` 
decrease in mixed treatments. Medium granules increased in relative abundance 
across all samples, while small granules decreased in wheat treatments and 
increased in potato treatments 
(Figure \@ref(fig:ratio-plots)).

@@

<!-- Big summary/important statement for the science -->
Here, we have provided a method for exploring the incorporation of dietary 
starches into the mineral matrix of a dental calculus biofilm model. Our results show
that a very low proportion of the starches exposed to the biofilm during growth are
retained in the mineral matrix, and that the size of the starch granules
may affect the likelihood of incorporation. The proportions of starch granules 
(of all sizes) present in the extracted samples were similar across all treatments
(`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`), 
despite large differences in absolute granule counts between wheat 
(mean = `r format(sol_comb_out$total[5], big.mark = ",")`) 
and potato
(mean = `r format(sol_comb_out$total[4], big.mark = ",")`)
solutions.  
The absolute counts, however, differed more visibly between treatments and was
proportional with the total count of granules in the treatment solutions. Wheat 
and mixed solutions had the highest absolute mean count of starch granules, and 
also had the highest absolute mean count of starch granules extracted from the 
dental calculus
(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
This suggests that the starches that are more frequently consumed will be present
in higher quantities in the dental calculus, at least prior to inhumation and 
degradation in the burial environment. 
Despite the low proportion of granules recovered from the model calculus 
(`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`),
the absolute counts were still substantially greater than counts recovered from 
archaeological remains 
[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossilContent2010], which could in part be due to the 
lack of oral amylase activity in our model.
